Software for the analysis and visualization of deep mutational scanning data

### 2015-03-23 paper/BMC_Bioinformatics_2015/ResponseToReviewers.tex

Bloom describes a software package for analysis of deep mutational scanning using a novel (for this application) Bayesian/MCMC framework. The general framework appears to be sound and addresses a significant need for these new and highly promising functional assays. The manuscript is well-written and explains the major uses cases of the software in sufficient detail.

Minor comments/suggestions:

1. The notion that differences based on larger absolute counts provide more certainty than those based on small absolute counts is obviously true, but only up to a point. Since deep mutational scanning experiments typically involves exponential amplification of a finite number of selected molecules prior to sequencing, there will be a saturation point (or region) beyond which higher absolute counts no longer provide additional information. This effect might not be negligible in experiments that involve typical cultures of $10^6$ to $10^8$ independently transformed or transduced cells. I would encourage the author to note these effects and perhaps add features or provide guidance to software users for how to estimate or detect the saturation point.

\begin{response}
This is an excellent point, and is completely true! The text has now been revised to emphasize that increasing deep-sequencing counts stop being meaningful once the typical number of counts per site becomes on the order of the number of unique molecules that were originally in the experiment. For most shotgun sequencing experiments, the typical counts per site ($\sim 10^5$ to $10^6$) still tend to be less than the typical number of unique molecules ($\sim 10^6$ to $10^8$), but this will not always be true particularly as sequencing depths increase.

The revised text now emphasizes this point. It emphasizes that the experiments should be done in a way to ensure that the sequencing depth is not saturating the number of unique molecules (i.e. by quantifying the number of molecules carefully). It also suggests that one way that this issue can be examined during \textit{post hoc} computational analyses is by sub-sampling the deep sequencing data (i.e. randomly selecting only a portion of the sequencing reads) when generated the mutation counts files used as input to inference programs in \textsl{dms\_tools}, and then seeing if this reduced depth alters the results. When the results start to become insensitive to sub-sampling the data, this provides an indication that the unique molecules are starting to be oversampled. 
\end{response}

@@

Question 1:
On page 2, in the second paragraph, a deep mutational scanning study is described. The average mutation rate is $\mu=1/L$, where $L$ is the length of the gene. It is unclear to me why this is so -- intuitively one might think that a longer gene may have more mutations, given that mutation is a random process. More generally, could the author provide more details on the range of data one would expect from typical deep mutational scanning experiments?

\begin{response}
In principle, a deep mutational scanning experiment could introduce any number of mutations per gene. But typically, experimentalists tend to introduce an average of $\approx$1 mutation per gene when they create the codon-mutant libraries. The reason is that having genes with $\gg$1 mutations makes it hard to attribute the selection on any given gene variant to any specific mutation. For this reason, published deep mutational scanning experiments have generally targeted $\overline{\mu} \sim 1/L$. This is an important distinction between deep mutational scanning and naturally occurring mutations -- naturally occurring mutations will be more common in longer genes, since typically the per-site mutation rate is approximately uniform and longer genes have more sites. The revised text now clarifies this point.
\end{response}

Question 2:
The author states that for a depth of $10^6$, we expect about 2000 counts of non-wild type codons for the average gene. The author arrived at this number by multiplying the total number of reads by the probability of a mutation. This is reasonable if each read comes from a different DNA fragment, each with probability $\mu$ of mutating. However, if multiple reads all cover the same DNA fragment, would this relationship still hold? The author should comment on whether this situation is likely to occur in practice and, if so, how it affects the method.

\begin{response}
This is an excellent comment, and is closely related to point (1) made by the first reviewer. The calculation to of the counts to which the reviewer refers, and indeed much of the statistical framework described in the manuscript, is made under the assumption that the number of unique DNA fragments in the library substantially exceeds the number of sequencing reads, such that the deep-sequencing reads can be viewed as subsampling a very large number of DNA fragments. This issue has been clarified in the text as described in the response to point (1) made by the first reviewer.
\end{response}

### 2015-03-23 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Responded to reviewer 1 comment 1, reviewer 2 comment 2

>>>>

The rest of this paper assumes that the sequencing depth is less than the number of unique molecules in the mutant library, such that the deep sequencing randomly subsamples the set of molecules. If this assumption is false (i.e. if the number of unique molecules is substantially less than the sequencing depth), then the accuracy of inferences about mutational effects will be fundamentally limited by this aspect of the experimental design. Properly done experiments should quantify the number of unique molecules in the library so that it is obvious whether this assumption holds. In the absence of such information, the analysis can be repeated using only a random fraction of the deep sequencing data to assess whether inferences are limited by sequencing depth or they underlying molecular diversity in the mutant library.
